---
title: 'Gene set enrichment analysis on hypergraph results'
author: |
  | Song Feng
  | Biological Sciences Division, Pacific Northwest National Laboratory
  | song.feng@pnnl.gov
date: "2020-05-26"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---

---

# Gene enrichment analysis of hypergraph results
From the hypergraph analysis, we have several lists of genes, now we want to check whethere these genes are enriched in some particular function groups/pathways. 

Firt, we get the gene lists from the excel files and txt files. Then use the gene set from hypergraph result as target list and use the gene list of particular function as functional sets/pathways and do the Fisher's exact test. The backgroud is all genes in bigTrans. There are three functional sets/pathways: interferon stimulated gene set (isgs), immune response set (ir), pathogen targeted gene list (pt).

```{r}
#install.packages("rlist")
library(rlist)
library(ggplot2)
library(latex2exp)

setwd("~/Dropbox/PNNL/Omics/SARS-COV-2/SARSCov2/")
# get the background gene set/list
biggerTransGeneSet <- as.character(read.table("biggerTransGenes.txt",  header = FALSE, sep = "")$V1)
lenBakgrd <- length(biggerTranGeneSet)
# get gene lists from functional sets
isgsGeneSet <- read.table("ISGs.txt", header = TRUE, sep = "\t")
isgsGeneList <- as.character(isgsGeneSet$Gene.Name)
irGeneList <- as.character(read.table("immuneResponse.txt", header = FALSE, sep = "")$V1)
ptGeneList <- as.character(unlist(read.table("pathogen_targets.gmt")))[-(1:4)]

source("GeneSets.R")
```

## Functions for calculating the enrichment scores (using p = 0, which is equavalent to Kolomogrov-Smirnov statistics)

```{r}
esFn <- function(x, p = 1) {
  len = length(x)
  miss = ifelse(x!=0, 0, 1)
  #hiti = ((miss * 2 - 1) - miss) / 2
  if (p == 0) {
    xp <- ifelse(x!=0, 1, 0)
  } else {
    xp = abs(x)**p
  }
  sumxp = sum(xp)
  summiss = sum(miss)
  if (sumxp == 0) {
    hitp = rep(0, len)
  } else {
    hitp = cumsum(xp) / sumxp
  }
  if (summiss == 0) {
    missp = rep(0, len)
  } else {
    missp = cumsum(miss) / summiss
  }
  maximum = max(hitp - missp)
  minimum = min(hitp - missp)
  score = ifelse(abs(maximum) > abs(minimum), maximum, minimum)
  return(score)
}

library(MASS)
epFn <- function(x, xnull) {
  fit <- fitdistr(xnull[xnull > 0], densfun="normal")
  pvalue = 1 - pnorm(abs((x - fit$estimate[1]) / (fit$estimate[2])))
  return(pvalue)
}
```



## Calculating the enrichment scores with different measurements

```{r}
samplesize = 100000
ss <- c(1:50)
genesetNames <- c("ISGS", "IR", "PT")
ztypes <- c("s-closeness", "s-betweenness")
#tsRankedGeneAllUniLen <- list()
dfAll <- setNames(data.frame(matrix(ncol = 7, nrow = 0)), c("genecount", "p", "Enrichment score", "p-value", "s", "Gene sets","Graph metrics"))
dfi <- 1
```

### Now analyse the s centrality and s betweenness scores

#### Using the average values of individual metrics

```{r}
for (ztype in ztypes) {
  infile <- sprintf("./biggerTrans-%s.csv", ztype)
  if (file.exists(infile)) {
    genelist <- read.csv(infile, header = TRUE, row.names = 1)
    allGenenames <- rownames(genelist)
    lenAll <- length(allGenenames)
    #sRankedGeneAll <- list()
    for (j in seq(1,length(ss))) {
      higherOrder <- as.matrix(genelist[1:j])
      aveHigherOrder <- sort(rowMeans(higherOrder), decreasing = TRUE)
      allZscores <- aveHigherOrder
      bakgrdZscores <- c(allZscores, rep(0, lenBakgrd - lenAll))
      aveHigherOrderNoZeros <- aveHigherOrder[aveHigherOrder!=0]
      genename <- names(aveHigherOrderNoZeros)
      zscore <- aveHigherOrderNoZeros
      dfsorted <- data.frame(genename, zscore)
      rownames(dfsorted) <- 1:nrow(dfsorted)
      ## calculate the statistics
      for (gs in 1:length(genesetNames)) {
        if (gs == 1) {
          functionSet = isgsGeneList
        } else if (gs == 2) {
          functionSet = irGeneList
        } else {
          functionSet = ptGeneList
        }
        # Fisher's exact test
        enr <- enrichment_by_fishers(dfsorted$genename, biggerTransGeneSet, functionSet)
        p = enr$fisher$p.value
        f = enr$foldx
        mat = enr$mat
        # ES calculation
        genenames = dfsorted$genename
        lenShort <- length(genenames)
        zscores <- dfsorted$zscore
        zscores[which(!(genenames %in% functionSet))] <- 0
        es <- esFn(zscores)
        sampledValues = c()
        for (k in 1:samplesize) {
          sampledis = sample(1:lenBakgrd)
          # tempGenenames <- biggerTransGeneSet[sampledis[1:lenShort]]
          # tempZscores <- bakgrdZscores[sampledis[1:lenShort]]
          tempGenenames <- genenames[sampledis[1:lenShort]]
          tempZscores <- zscores[sampledis[1:lenShort]]
          tempZscores[which(!(tempGenenames %in% functionSet))] <- 0
          sampledValues = c(sampledValues, esFn(tempZscores))
        }
        
        ep = (length(which(sampledValues >= es)) + 1) / (samplesize + 1)
        #ep = epFn(es, sampledValues)
        # save test result
        dfAll[dfi,] <- c(mat[1,1], p, es, ep, ss[j], genesetNames[gs], ztype)
        dfi <- dfi + 1
      }
    }
    #tsRankedGeneAllUniLen <- c(tsRankedGeneAllUniLen, list(sRankedGeneAll))
  }
}
#save(tsRankedGeneAllUniLen, file = "tsRankedGeneAllUniLen.RData")

```


### Make all the plotable values as numeric and save to file

```{r}
save(dfAll, file = "dfAllAve.RData")
```

## Load and plot results

```{r}
load("dfAllAve.RData")
dfAll$genecount <- as.numeric(dfAll$genecount)
dfAll$p <- as.numeric(dfAll$p)
dfAll$`Enrichment score` <- as.numeric(dfAll$`Enrichment score`)
dfAll$`p-value` <- as.numeric(dfAll$`p-value`)
dfAll$s <- as.numeric(dfAll$s)
```

Updates@2020-01-13 Generate plots of s <= 45 (also for p value plot) with z threshold = 3 and with p value threshold

```{r}
df <- dfAll
ggplot(data = df, mapping = aes(x=s, y=`Enrichment score`, color = `Gene sets`, shape = `Graph metrics`)) + geom_point(size=3 ,alpha = 0.6) + geom_line(linetype = "dashed") + scale_x_continuous(name='s', limits=c(0, 50)) + scale_y_continuous(limits = c(0.0, NA))
```


```{r}
ggplot(data = df, mapping = aes(x=s, y=`p-value`, color = `Gene sets`, shape = `Graph metrics`)) + geom_point(size=3 ,alpha = 0.6) + geom_line(linetype = "dashed") + scale_x_continuous(name=TeX("s"),  limits=c(0, 50)) + scale_y_log10(limits = c(NA, 1))
```
## Calculating the enrichment scores with different measurements

```{r}
samplesize = 100000
ss <- c(1:50)
genesetNames <- c("ISGS", "IR", "PT")
ztypes <- c("s-closeness", "s-betweenness")
#tsRankedGeneAllUniLen <- list()
dfNoSars2 <- setNames(data.frame(matrix(ncol = 7, nrow = 0)), c("genecount", "p", "Enrichment score", "p-value", "s", "Gene sets","Graph metrics"))
dfi <- 1
```

### Now analyse the s centrality and s betweenness scores

#### Using the average values of individual metrics

```{r}
for (ztype in ztypes) {
  infile <- sprintf("./biggerTransNoSars2-%s.csv", ztype)
  if (file.exists(infile)) {
    genelist <- read.csv(infile, header = TRUE, row.names = 1)
    allGenenames <- rownames(genelist)
    lenAll <- length(allGenenames)
    #sRankedGeneAll <- list()
    for (j in seq(1,length(ss))) {
      higherOrder <- as.matrix(genelist[1:j])
      aveHigherOrder <- sort(rowMeans(higherOrder), decreasing = TRUE)
      allZscores <- aveHigherOrder
      bakgrdZscores <- c(allZscores, rep(0, lenBakgrd - lenAll))
      aveHigherOrderNoZeros <- aveHigherOrder[aveHigherOrder!=0]
      genename <- names(aveHigherOrderNoZeros)
      zscore <- aveHigherOrderNoZeros
      dfsorted <- data.frame(genename, zscore)
      rownames(dfsorted) <- 1:nrow(dfsorted)
      ## calculate the statistics
      for (gs in 1:length(genesetNames)) {
        if (gs == 1) {
          functionSet = isgsGeneList
        } else if (gs == 2) {
          functionSet = irGeneList
        } else {
          functionSet = ptGeneList
        }
        # Fisher's exact test
        enr <- enrichment_by_fishers(dfsorted$genename, biggerTransGeneSet, functionSet)
        p = enr$fisher$p.value
        f = enr$foldx
        mat = enr$mat
        # ES calculation
        genenames = dfsorted$genename
        lenShort <- length(genenames)
        zscores <- dfsorted$zscore
        zscores[which(!(genenames %in% functionSet))] <- 0
        es <- esFn(zscores)
        sampledValues = c()
        for (k in 1:samplesize) {
          sampledis = sample(1:lenBakgrd)
          # tempGenenames <- biggerTransGeneSet[sampledis[1:lenShort]]
          # tempZscores <- bakgrdZscores[sampledis[1:lenShort]]
          tempGenenames <- genenames[sampledis[1:lenShort]]
          tempZscores <- zscores[sampledis[1:lenShort]]
          tempZscores[which(!(tempGenenames %in% functionSet))] <- 0
          sampledValues = c(sampledValues, esFn(tempZscores))
        }
        
        ep = (length(which(sampledValues >= es)) + 1) / (samplesize + 1)
        #ep = epFn(es, sampledValues)
        # save test result
        dfNoSars2[dfi,] <- c(mat[1,1], p, es, ep, ss[j], genesetNames[gs], ztype)
        dfi <- dfi + 1
      }
    }
    #tsRankedGeneAllUniLen <- c(tsRankedGeneAllUniLen, list(sRankedGeneAll))
  }
}
#save(tsRankedGeneAllUniLen, file = "tsRankedGeneAllUniLen.RData")

```


### Make all the plotable values as numeric and save to file

```{r}
save(dfNoSars2, file = "dfNoSars2Ave.RData")
```

## Load and plot results

```{r}
load("dfNoSars2Ave.RData")
dfAll$genecount <- as.numeric(dfAll$genecount)
dfAll$p <- as.numeric(dfAll$p)
dfAll$`Enrichment score` <- as.numeric(dfAll$`Enrichment score`)
dfAll$`p-value` <- as.numeric(dfAll$`p-value`)
dfAll$s <- as.numeric(dfAll$s)
```

Updates@2020-01-13 Generate plots of s <= 45 (also for p value plot) with z threshold = 3 and with p value threshold

```{r}
df <- dfNoSars2
ggplot(data = df, mapping = aes(x=s, y=`Enrichment score`, color = `Gene sets`, shape = `Graph metrics`)) + geom_point(size=3 ,alpha = 0.6) + geom_line(linetype = "dashed") + scale_x_continuous(name='s', limits=c(0, 50)) + scale_y_continuous(limits = c(0.0, NA))
```


```{r}
ggplot(data = df, mapping = aes(x=s, y=`p-value`, color = `Gene sets`, shape = `Graph metrics`)) + geom_point(size=3 ,alpha = 0.6) + geom_line(linetype = "dashed") + scale_x_continuous(name=TeX("s"),  limits=c(0, 50)) + scale_y_log10(limits = c(NA, 1))
```

### Get all the ranks based on the average and also 

```{r}
for (ztype in ztypes) {
  infile <- sprintf("./biggerTrans-%s.csv", ztype)
  outfile <- sprintf("./biggerTransRanks-%s.csv", ztype)
  if (file.exists(infile)) {
    genelist <- read.csv(infile, header = TRUE, row.names = 1)
    allGenenames <- rownames(genelist)
    lenAll <- length(allGenenames)
    firstCol <- as.matrix(genelist[1])
    ranked <- data.frame(rank = seq(1:length(firstCol[firstCol!=0])))
    rownames(ranked) <- ranked$rank
    for (j in seq(1,length(ss))) {
      higherOrder <- as.matrix(genelist[1:j])
      aveHigherOrder <- sort(rowMeans(higherOrder), decreasing = TRUE)
      allZscores <- aveHigherOrder
      bakgrdZscores <- c(allZscores, rep(0, lenBakgrd - lenAll))
      aveHigherOrderNoZeros <- aveHigherOrder[aveHigherOrder!=0]
      genename <- names(aveHigherOrderNoZeros)
      zscore <- aveHigherOrderNoZeros
      dfsorted <- data.frame(genename, zscore)
      rownames(dfsorted) <- 1:nrow(dfsorted)
      
      ranked <- cbind(ranked, dfsorted)
    }
    write.csv(ranked, outfile)
  }
}

```


```{r}
for (ztype in ztypes) {
  infile <- sprintf("./biggerTransNoSars2-%s.csv", ztype)
  outfile <- sprintf("./biggerTransNoSars2Ranks-%s.csv", ztype)
  if (file.exists(infile)) {
    genelist <- read.csv(infile, header = TRUE, row.names = 1)
    allGenenames <- rownames(genelist)
    lenAll <- length(allGenenames)
    firstCol <- as.matrix(genelist[1])
    rankedNoSars2 <- data.frame(rank = seq(1:length(firstCol[firstCol!=0])))
    rownames(ranked) <- ranked$rank
    for (j in seq(1,length(ss))) {
      higherOrder <- as.matrix(genelist[1:j])
      aveHigherOrder <- sort(rowMeans(higherOrder), decreasing = TRUE)
      allZscores <- aveHigherOrder
      bakgrdZscores <- c(allZscores, rep(0, lenBakgrd - lenAll))
      aveHigherOrderNoZeros <- aveHigherOrder[aveHigherOrder!=0]
      genename <- names(aveHigherOrderNoZeros)
      zscore <- aveHigherOrderNoZeros
      dfsorted <- data.frame(genename, zscore)
      rownames(dfsorted) <- 1:nrow(dfsorted)
      
      rankedNoSars2 <- cbind(rankedNoSars2, dfsorted)
    }
    write.csv(rankedNoSars2, outfile)
  }
}

```


## Analyze the ranked lists

```{r}
source("bagOfGenesEnrichment.R")
```

```{r}
metrics <- c("sbetweenness", "scloseness")
ranks <- c("relValChange", "relRankChange", "euclidean", "numIncreases")

for (metric in metrics) {
  for (rank in ranks) {
    inFile <- paste0("rankedLists/",metric, "_", ranks, ".txt")
    
  }
}



```

